\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-05 submission/manuscript.Rmd
collinearity:

@@

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** (A) L2 logistic regression coefficients  (B) L1 SVM with linear kernel feature weights (C) L2 SVM with linear kernel feature ranks. The feature ranks of the most important five OTUs in each data-split are shown here. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The training times for training and testing of each data-split showed the differences in computational efficieny of the seven models. The median traintime in hours was the highest for XGBoost and shortest for L2 logistic regression. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage
\includegraphics{Figure_S1.png}
**Figure S1. Hyperparameter setting performances for linear models.** (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters are used in training the model. The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.

\newpage
\includegraphics{Figure_S2.png}
**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model. The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
\newpage

\includegraphics{TableS1.pdf}
**Table S1. ML model complexity and hyperparameter range.** 
\newpage

>>>>

We also used permutation importance to interpret linear models to show the differences between the two interpretation methods [Figure S3]. L1-regularized SVM with linear kernel picked out the same OTUs as the most important features in methods that use feature rankings [Figure 3] as well as permutation importance [Figure S3]. However, L2-regularized models had different feature importances in the two methods of interpretation due to collinearity in microbial communities.

@@

ML model interpretation methods in the previous studies microbiome studies lacked clarity and transparency. The studies that made the model interpreatation code publicly available, used default functions in previously built packages in R, python and matlab. In this study we used two methods to interpret our models. Feature weights of trained models ad permutation importance. the differences between the two interpretation methods

@@

\newpage
\includegraphics{Figure_1}
**Figure 1. Machine learning pipeline showing predictive model training and evaluation flowchart.  **  We split the data 80%/20% stratified to maintain the overall label distribution, performed five-fold cross-validation on the training data to select the best hyperparameter setting and then using these hyperparameters to train all of the training data. The model was evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic curve
\newpage
\includegraphics{Figure_2.png}
**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross validation and testing performances. **   The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by horizontal line at 0.50) for all the ML models. Discriminative performance of random forest model was higher than other ML models. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: SRN, screen-relevant neoplasias; AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics{Figure_3.png}
**Figure 3. Interpretation of the linear ML models.** The absolute feature weights of (A) L2 logistic regression coefficients (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel were ranked from highest rank 1 to 100 for each data-split. The feature ranks of the highest ranked five OTUs based on their median ranks are shown here. Similar OTUs had the largest impact on the predictive performance of L2 logistic regression and L2 SVM with linear kernel. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_4.png}
**Figure 4. Explanation of the non-linear ML models.** (A) SVM with radial basis kernel (B) decision tree (C) random forest (D) XGboost feaure importances were explanied using permutation importance using held-out test set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed. The colors of the boxplots stand for the unique OTUs that are shared among the different models; pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for OTU00110, green for OTU00361 and red for OTU00882. For all the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance of the model. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png}
**Figure 5. Computational efficiency of seven ML models.** The training times for of each data-split showed the differences in computational efficieny of the seven models. The median training time in hours was the highest for XGBoost and shortest for L1-regularized SVM with linear kernel. The boxplot shows quartiles at the box ends and the statistical median as the horizontal line in the box. The whiskers show the farthest points that are not outliers. Outliers are data points that are not within 3/2 times the interquartile ranges. 
Abbreviations: AUROC, area under the receiver operating characteristic curve; SVM, support vector machine; XGBoost, extreme gradient boosting.  
\newpage
\includegraphics{Figure_S1.png}
**Figure S1. Hyperparameter setting performances for linear models.** (A) L2 logistic regression (B) L1 SVM with linear kernel (C) L2 SVM with linear kernel mean cross-validation AUROC values when different hyperparameters are used in training the model. The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.

\newpage
\includegraphics{Figure_S2.png}
**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree (B) Random forest (C) SVM with radial basis kernel (D) XGoost mean cross-validation AUROC values when different hyperparameters are used in training the model. The differences in AUROC values when hyperparameters change show that hyperparameter tuning is a crucial step in building a ML model.
\newpage
\includegraphics{Figure_S4.png}
**Figure S3. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel (B) L2-regularized SVM with linear kernel and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation performed. 
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\includegraphics{TableS1.pdf}
**Table S1. ML model complexity and hyperparameter range.** 
\newpage

### 2019-07-05 submission/manuscript.tex
collinearity:

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
ranks. The feature ranks of the most important five OTUs in each
data-split are shown here. Similar OTUs had the largest impact on the
predictive performance of L2 logistic regression and L2 SVM with linear
kernel. Abbreviations: SVM, support vector machine; OTU, Operational
Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for training and
testing of each data-split showed the differences in computational
efficieny of the seven models. The median traintime in hours was the
highest for XGBoost and shortest for L2 logistic regression. The boxplot
shows quartiles at the box ends and the statistical median as the
horizontal line in the box. The whiskers show the farthest points that
are not outliers. Outliers are data points that are not within 3/2 times
the interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2 logistic regression (B)
L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model.

@@

\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} (A) Decision tree (B)
Random forest (C) SVM with radial basis kernel (D) XGoost mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model. \newpage

\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
hyperparameter range.} \newpage

>>>>

We also used permutation importance to interpret linear models to show
the differences between the two interpretation methods {[}Figure S3{]}.
L1-regularized SVM with linear kernel picked out the same OTUs as the
most important features in methods that use feature rankings {[}Figure
3{]} as well as permutation importance {[}Figure S3{]}. However,
L2-regularized models had different feature importances in the two
methods of interpretation due to collinearity in microbial communities.

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} The absolute feature weights of (A) L2 logistic
regression coefficients (B) L1 SVM with linear kernel (C) L2 SVM with
linear kernel were ranked from highest rank 1 to 100 for each
data-split. The feature ranks of the highest ranked five OTUs based on
their median ranks are shown here. Similar OTUs had the largest impact
on the predictive performance of L2 logistic regression and L2 SVM with
linear kernel. Abbreviations: SVM, support vector machine; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. The colors of the boxplots
stand for the unique OTUs that are shared among the different models;
pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for
OTU00110, green for OTU00361 and red for OTU00882. For all the
tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
the largest impact on predictive performance of the model.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.

@@

\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for of each
data-split showed the differences in computational efficieny of the
seven models. The median training time in hours was the highest for
XGBoost and shortest for L1-regularized SVM with linear kernel. The
boxplot shows quartiles at the box ends and the statistical median as
the horizontal line in the box. The whiskers show the farthest points
that are not outliers. Outliers are data points that are not within 3/2
times the interquartile ranges. Abbreviations: AUROC, area under the
receiver operating characteristic curve; SVM, support vector machine;
XGBoost, extreme gradient boosting.\\
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2 logistic regression (B)
L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model.

\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} (A) Decision tree (B)
Random forest (C) SVM with radial basis kernel (D) XGoost mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model. \newpage
\includegraphics{Figure_S4.png} \textbf{Figure S3. Interpretation of the
linear ML models with permutation importance.} (A) L1-regularized SVM
with linear kernel (B) L2-regularized SVM with linear kernel and (C)
L2-regularized logistic regression were interpreted using permutation
importance using held-out test set. The gray rectangle and the dashed
line show the IQR range and median of the base testing AUROC without any
permutation performed. Abbreviations: SVM, support vector machine; OTU,
Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational
Taxonomic Unit. \includegraphics{TableS1.pdf} \textbf{Table S1. ML model
complexity and hyperparameter range.} \newpage
